Neural networks for solving regression and classification problems

### 2022-11-16 doc/paper.tex
figs, many figs

The Franke Function is a simple weighted sum of polynomials, taking two inputs and returning a real value. It is commonly used to test machine learning algorithms for regression problems. For our linear regression problems, our training data will be based on a 9th degree polynomial approximation to the Franke Function, with 55 features. For the feed forward neural network, we will use 2 of the features (x, y) of the Franke Function.

@@

\begin{table}[h]
\caption{Characteristics of each dataset used in the report}
\begin{center}
\label{datasettable}
\begin{tabular}{| c | l | l | l |}
\hline
Dataset & Number of instances & Features & Type \\
\hline
Franke Function & synthetic & 55 & regression \\
Franke Function & synthetic & 2 & regression \\
Terrain (downscaled) & 204 800 & 2 & regression \\
Wisconsin Breast Cancer & 569 & 30 & binary classification \\
MNIST 8x8 & 1797 & 64 & multiclass classification (10 classes) \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

From observing FIG CONSTANT V HESSIAN we see that Newton's method using Hessian inversion greatly outperforms the constant scheduler by reaching an MSE of 0.02 for the testing data within the first epoch. This is due to the Hessian inversion in the first epoch in Newton's method being equivalent to the analytical solution.  By comparison, the constant scheduler given 200 epochs reaches an MSE of 0.198 for the testing data. Given this result, it would seem indefensible to claim that Newtons method requires improvements, especially considering the time it took the Newton method 0.1512 seconds to compute 200 epochs, again beating the result from the constant scheduler which took 0.4991 seconds \footnote{this result should be taken with a grain of salt, given that the constant scheduler is run as a part of a larger architecture, whilst the Hessian calculation is done in a seperate script. Still, this provides an intuition that Newton's method does not take much time to compute this specific problem}. This result is due to the Hessian matrix only being a mere 55x55 matrix for this problem, such that the matrix inversion will not be computationally complex. However, if we were to increase the complexity of the problem, Newtons method would rapidly increase in computational complexity as it is a large matrix inversion with many floating point operations. For problems of such a small scale, when our features are based on a polynomial approximation Franke Function, it can be important to remember that we don't necessarily have to turn to these fancier solutions as the simple Newton method will do the job perfectly. Keeping this lesson in mind, we wish to further investigate the various scheduler methods in order to gain a better understanding of how well they perform, which will become useful for larger problems.

@@

@@

@@

Another observation from FIG NUM is that the schedulers final error for 100 epochs are more or less the same. As the largest change in MSE occurs in the first 50 epochs as seen in ALL FIGS, perhaps the other schedulers are given enough time to catch up as the best performing schedulers, such as Adagrad, starts slowing down the rate at which the MSE is decreased. However, as can be seen in HEATMAP FIG, Adagrad offers a 12\% \footnote{$\frac{constant MSE}{adagrad MSE} * 100 = \frac{0.018}{0.015} * 100 = 12\%$} decrease in MSE compared to the constant scheduler. It is clear from the results thus far that Adagrad is the best scheduler for approximating the Franke function when using linear regression with gradient descent. We believe that the reason Adagrad wins out is that these result are based on training for 100 epochs. The downfall of Adagrad in the literature seems to be that the aggressive decaying of the learning rate hinders a quick convergence deeper into training SOURCE https://towardsdatascience.com/learning-parameters-part-5-65a2f3583f7d. We are thus cautionary to claim that Adagrad will do the best job over a much larger amount of epochs when compared to adaptive schedulers that claim to improve this aspect.

To test this, we compared Adagrad with Adam over 5000 epochs. Both schedulers performed equally, both resulting in an MSE of 0.017 with Adagrad taking the edge with an MSE 0.0002 lower. Scikit obtained a test data MSE of 0.005, but followed the same trend in performance of dipping early and stabilizing after just a 100 epochs.

Furthermore, when trained for 5000 epochs, Adam with optimal parameters keeps improving the MSE. Scikit learn's Adam scheduler does perform better, offering a greater decrease the MSE. We compared Adagrad to scikit learn's Adam MLPRegressor as this seemed to be the closest option for comparing our results. Even though we have cross validated both our own Adagrad implementation and Scikit's with 5 folds, these differences in results might be due to the difference in solver, a more favorable weight initiation in Scikits model or additional optimizations present in the scikit learn implementation. Nonetheless, neither models are able to beat the analytical solution provided by linear regression methods using polynomial approximations, with our best result of an MSE of 0.012 on the test data and Scikits final MSE of 0.004 (granted this is on the train data, as scikit learn does not provide the loss curve for the test data).  both falling short of the analytical MSE of 0,003. In theory, the feed forward neural network should be able to better approximate the Franke function given that it is not limited to polynomial solutions. This result might be due to 5000 epochs not being enough training time, as the Franke function is a quite complex function with added noise.

>>>>

The Franke Function is a simple weighted sum of polynomials, taking two inputs and returning a real value. It is commonly used to test machine learning algorithms for regression problems. For our linear regression problems, our training data will be based on a 9th degree polynomial approximation to the Franke Function, with 56 features. For the feed forward neural network, we will use 2 of the features (x, y) of the Franke Function.

@@

\begin{table}[h]
\caption{Characteristics of each dataset used in the report}
\begin{center}
\label{datasettable}
\begin{tabular}{| c | l | l | l |}
\hline
Dataset & Number of instances & Features & Type \\
\hline
Franke Function & synthetic & 56 & regression \\
Franke Function & synthetic & 2 & regression \\
Terrain (downscaled) & 204 800 & 2 & regression \\
Wisconsin Breast Cancer & 569 & 30 & binary classification \\
MNIST 8x8 & 1797 & 64 & multiclass classification (10 classes) \\
\hline
\end{tabular}
\end{center}
\end{table}

@@

From observing FIG CONSTANT V HESSIAN we see that Newton's method using Hessian inversion greatly outperforms the constant scheduler by reaching an MSE of 0.02 for the testing data within the first epoch. This is due to the Hessian inversion in the first epoch in Newton's method being equivalent to the analytical solution.  By comparison, the constant scheduler given 200 epochs reaches an MSE of 0.198 for the testing data. Given this result, it would seem indefensible to claim that Newtons method requires improvements, especially considering the time it took the Newton method 0.1512 seconds to compute 200 epochs, again beating the result from the constant scheduler which took 0.4991 seconds \footnote{this result should be taken with a grain of salt, given that the constant scheduler is run as a part of a larger architecture, whilst the Hessian calculation is done in a seperate script. Still, this provides an intuition that Newton's method does not take much time to compute this specific problem}. This result is due to the Hessian matrix only being a mere 56x56 matrix for this problem, such that the matrix inversion will not be computationally complex. However, if we were to increase the complexity of the problem, Newtons method would rapidly increase in computational complexity as it is a large matrix inversion with many floating point operations. For problems of such a small scale, when our features are based on a polynomial approximation Franke Function, it can be important to remember that we don't necessarily have to turn to these fancier solutions as the simple Newton method will do the job perfectly. Keeping this lesson in mind, we wish to further investigate the various scheduler methods in order to gain a better understanding of how well they perform, which will become useful for larger problems.

@@

From FIGURE MOMENTUM  v CONSTANT, we can observe that adding momentum does not improve upon the solutions found by the constant scheduler when using full gradient descent, reaching once again an MSE of 0.21 for the test data. This implies that when both schedulers are gridsearched for optimal hyperparameters, the optimal value for the momentum hyperparameter is 0. This result makes sense, as the momentum hyperparameter is used to reduce noise from derivatives calculated from smaller batch sizes, so-called mini batches. When utilizing a full gradient descent, where the batch size equals the size of the training data, momentum will not improve our results.

@@

By introducing mini-batch gradient descent, we can observe that

@@

Another observation from FIG NUM is that the schedulers final error for 100 epochs are more or less the same. As the largest change in MSE occurs in the first 50 epochs as seen in ALL FIGS, perhaps the other schedulers are given enough time to catch up as the best performing schedulers, such as Adagrad, starts slowing down the rate at which the MSE is decreased. However, as can be seen in HEATMAP FIG, Adagrad offers a 12\% \footnote{$\frac{constant MSE}{adagrad MSE} * 100 = \frac{0.018}{0.015} * 100 = 12\%$} decrease in MSE compared to the constant scheduler. It is clear from the results thus far that Adagrad is the best scheduler for approximating the Franke function when using linear regression with gradient descent. We believe that the reason Adagrad wins out is that these result are based on training for 100 epochs. The downfall of Adagrad in the literature seems to be that the aggressive decaying of the learning rate hinders a quick convergence deeper into training SOURCE https://towardsdatascience.com/learning-parameters-part-5-65a2f3583f7d. We are thus cautionary to claim that Adagrad will do the best job over a much larger amount of epochs when compared to adaptive schedulers that claim to improve this aspect such as Adam.

5000e_adam_v_adagrad

To test this, we compared Adagrad with Adam over 5000 epochs, cross validated for 5 folds. Both schedulers perform almost equally, with Adagrad reaching a slightly lower MSE of 0.011 for the test data when compared to Adam's 0.012. It could be argued that 5000 epochs is not enough for the improvements of Adam to be noticeable, as the MSE could still be pushed much lower. However, for these computationally expensive runs it is clear that Adagrad is the scheduler of choice, especially considering our previous conclusion that Adagrad used less time to train. It should be noted that scikit learn's implementation outperforms both our implementation of Adagrad and Adam, with an MSE of 0.004. This could be due to a more favorable wight initialization in scikits model, along with additional optimizations implemented in their library. That we don't beat the analytical solution is no surprise, as we have limited ourself to a polynomial approximation of the Franke Function with 56 features. Our MSE might approach the analytical MSE, but this might take an unreasonable amount of epochs considering the computational runtime. Next, we investigate the use of linear regression in a neural network with hidden nodes, such that we can free ourselves of the limitations of the polynomial approximation.
